\documentclass[12pt,a4paper]{article}

\usepackage{graphicx}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage[margin=1in]{geometry} % Optional: tighter margins for A4 fit
\usepackage{bbm}
\usepackage{float}
\usepackage{array}
\setcounter{section}{-1}


\title{RL FYP}
\author{Mykola Vaskevych}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
something abstract
\end{abstract}


\section{random}

geeks rl:\cite{GeeksforGeeks2018RL}


\section{summary chapters 1-6 from rl book}
from book \cite{Sutton2018}



\subsection{Chapter 1}


RL is different from supervised and unsupervised learning. It tries to maximize the "reward signal" rather than finding hidden structure. RL is a balance between exploring new ways to obtain rewards and exploiting known methods.
RL has (from webpage' geeks for geeks' \cite{GeeksforGeeks2018RL}):  

    \begin{itemize}
        \item Agent: The decision-maker who performs actions.
        \item Environment: The world or system in which the agent operates.
        \item State: The situation or condition in which the agent is currently situated.
        \item Action: The possible moves or decisions that the agent can make.
        \item Reward: The feedback or result from the environment based on the agent's action.
    \end{itemize} 

    
The hardest part is estimating the values that reinforcement learning (RL) uses to obtain a reward, as these values need to be recalculated repeatedly, while something like the reward is simply provided by the environment. Thus, this is the most important aspect of any reinforcement learning (RL)


RL can be similar to evolutionary algorithms in some cases, but it is usually better, while evolutionary algorithms do not notice why certain samples "survived" and moved to the next generation, RL utilizes policy knowledge and a reward system to understand exactly why certain decisions are better.


\subsection{summary of ch1 rephrased}
Reinforcement Learning is a Machine Learning approach that teaches machines to understand and automate goal-directed learning and decision-making in specific tasks, such as optimizing data center management or playing the tic-tac-toe game. Its main difference from Supervised or unsupervised learning is that RL does not require supervision or a complete environment model. It's the first ML approach that learns from interactions with the environment to achieve long-term goals.

RL defines the interaction between an agent and the environment and is based on a framework of Markov decision processes. Its main components are: Agent, Environment, State, Action, and Reward. The concept of value and the key value function is the main concept for many RL algorithms, including those that we will discuss later, value functions are important for efficient search in the space of policies, and they are the main difference that separates evolutionary algorithms from RL, since evolutionary algorithms search directly in the policy space.




\subsection{Chapter 2}
The difference from other methods unique to RL is that we give a score to the actions taken rather than saying correct/incorrect.
In RL, you need to balance between exploitation of already learned actions with good reward and exploration to potentially find even better actions to exploit for better reward in the long-term goal.
\begin{itemize}
    \item Exploitation: Playing the machine that seems best so far
    \item Exploration: Trying other machines to see if they're actually better
\end{itemize}



\subsubsection{K-armed bandit problem}


\begin{enumerate}

\item \textbf{Sample-Average Method}
\begin{itemize}
    \item Estimates action value by getting the average of all previous rewards
    \item $$Q_t(a) = \frac{\sum_{i=1}^{t-1} R_i \cdot \mathbbm{1}_{A_i=a}}{\sum_{i=1}^{t-1} \mathbbm{1}_{A_i=a}}$$
    \item Simple but slow to adapt
\end{itemize}

\item \textbf{Greedy Action Selection}
\begin{itemize}
    \item Always selects action with highest estimated value
    \item $$A_t = \arg\max_a Q_t(a)$$
    \item Pure exploitation and no exploration can get stuck on suboptimal actions
\end{itemize}

\item \textbf{$\varepsilon$-Greedy Method}
\begin{itemize}
    \item Selects best action with probability $1-\varepsilon$, random action with probability $\varepsilon$
    \item $$A_t = \begin{cases} \arg\max_a Q_t(a) & \text{prob } 1-\varepsilon \\ \text{random} & \text{prob } \varepsilon \end{cases}$$
    \item Balances exploration and exploitation, ensures all actions sampled infinitely
\end{itemize}

\item \textbf{Incremental Implementation}
\begin{itemize}
    \item Efficient update without storing all rewards
    \item $$Q_{n+1} = Q_n + \frac{1}{n}[R_n - Q_n]$$
    \item General form: 
    $$\text{NewEstimate} \leftarrow \text{OldEstimate} + \text{StepSize}[\text{Target} - \text{OldEstimate}]$$
    \item Constant memory and computation per step
\end{itemize}

\item \textbf{Constant Step-Size Method}
\begin{itemize}
    \item Uses fixed step-size for nonstationary problems
    \item $$Q_{n+1} = Q_n + \alpha[R_n - Q_n], \quad \alpha \in (0,1]$$
    \item Exponential recency-weighted average: 
    $$Q_{n+1} = (1-\alpha)^n Q_1 + \sum_{i=1}^{n} \alpha(1-\alpha)^{n-i} R_i$$
    \item Weights recent rewards more; doesn't fully converge but adapts to changes
    \item Convergence conditions: 
    $$\sum_{n=1}^{\infty} \alpha_n(a) = \infty \quad \text{and} \quad \sum_{n=1}^{\infty} \alpha_n^2(a) < \infty$$
\end{itemize}

\item \textbf{Optimistic Initial Values}
\begin{itemize}
    \item Initialize all $Q_1(a)$ to high values (e.g., $+5$)
    \item No new formula, uses standard updates with optimistic initialization
    \item Encourages exploration through "disappointment", temporary effect, not good for problems that change over time
\end{itemize}

\item \textbf{Upper Confidence Bound (UCB)}
\begin{itemize}
    \item Selects options with the highest potential to be best
    \item $$A_t = \arg\max_a \left[Q_t(a) + c\sqrt{\frac{\ln t}{N_t(a)}}\right]$$
    \item Where $N_t(a)$ = times action $a$ selected, $c > 0$ controls exploration
    \item Deterministic exploration, favors uncertain actions, difficult to extend beyond bandits problems type
\end{itemize}

\item \textbf{Gradient Bandit Algorithm}
\begin{itemize}
    \item Learns preferences $H_t(a)$, not values
    \item Soft-max distribution: 
    $$\pi_t(a) = \frac{e^{H_t(a)}}{\sum_{b=1}^{k} e^{H_t(b)}}$$
    \item Update rule: 
    $$H_{t+1}(a) = H_t(a) + \alpha(R_t - \bar{R}_t)(\mathbbm{1}_{a=A_t} - \pi_t(a))$$
    \item Baseline $\bar{R}_t$ is average reward, stochastic gradient ascent method
    \item Performance gradient: 
    $$\frac{\partial \mathbb{E}[R_t]}{\partial H_t(a)} = \mathbb{E}[(R_t - \bar{R}_t)(\mathbbm{1}_{a=A_t} - \pi_t(a))]$$
\end{itemize}

\item \textbf{Associative Search (Contextual Bandits)}
\begin{itemize}
    \item Learn policy mapping situations to actions
    \item $$\pi: \mathcal{S} \to \mathcal{A}$$
    \item Bridge between bandits and full RL; actions don't affect next situation
    \item True action value: 
    $$q_*(a) = \mathbb{E}[R_t \mid A_t = a]$$
    \item Expected reward: 
    $$\mathbb{E}[R_t] = \sum_x \pi_t(x) q_*(x)$$
\end{itemize}

\end{enumerate}

\subsubsection{comparison}


\begin{table}[H]
\centering
\caption{Comparison of K-Armed Bandit Algorithms}
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Algorithm} & \textbf{Exploration Type} & \textbf{Adapts to Changes} & \textbf{Complexity} \\ \hline
Greedy & None & No & Simplest \\ \hline
$\varepsilon$-greedy & Random & With constant $\alpha$ & Simple \\ \hline
Optimistic Init & Early only & No & Simple \\ \hline
UCB & Intelligent & Difficult & Medium \\ \hline
Gradient Bandit & Probabilistic & Yes & Medium \\ \hline
\end{tabular}
\end{table}

\subsubsection{summary of chapter 2}
This chapter explains some of the methods for deciding which actions should be taken. Although these methods are considered basic and simple by the book, they are still useful. The book states that there are more advanced methods. However, they are much more complex and are not covered in this book. 

From graphs that I haven't included here, we can see that all algos introduced work the best when values in formulas are somewhere in the middle all are u shaped, but UCB so far the best performer


\subsection{Chapter 3}
Introduced Markov's decision processes, or MDPs. It's an idealized version of Reinforcment learning where brute force math can be applied, basically its a way how to formulate desigion making sequence, where sequence means not only action taken now, but in current situation. this itroduces to us delayed reward and state.


MDP math proof tells us that any problem that has a clear goal can be reduced to three signals that go back and forth between the agent and the environment: one for the state, one for the agent, and one for the reward.

usage of reward signal is the most certain difference of RL in comparison to other methods of ML.

policies are hard to compute; they require powerful hardware and a lot of time, so most of the time we need to do an approximation.

Terms like policy, value function, reward and two types of tasks: episodic and continuing were introduced and explained. later they were combined in one formula: Bellman optimality equation, which allows us to define optimal values, without having optimal policy first.





\subsection{Chapter 4}
Dynamic Programming is a collection of algorithms that compute optimal policies when a complete environment is given.  For DP to work you need:a finite MDP, all states, all rewards, and all possible actions as well as all dynamics ($p(s',r\mid s,a)$)

instead of solving Bellman equation iterate:

$\text{Old value} \Rightarrow \text{Update rule} \Rightarrow \text{New value} \Rightarrow \text{Repeat until merge}$

Two methods are described: policy iteration and value iteration.
\begin{itemize}
    \item Policy Iteration: less iterations, more expensive updates
    \item Value Iteration: more iterations, cheap updates
    \item both converge fast in practice.
\end{itemize}



\subsection{Chapter 5}

Monte Carlo (MC) methods learn from experience without needing a model of the environment. Unlike DP, which needs all dynamics $p(s',r|s,a)$, MC learns from experience. MC works only for episodic tasks (games that end).

\subsubsection{Basic Idea}
To estimate $v_\pi(s)$: follow policy $\pi$, play many episodes, every time you visit state $s$ record the return $G_t$, then average all returns. Law of large numbers guarantees convergence to true value.

Two variants exist:
\begin{itemize}
    \item First-Visit MC: only count first time $s$ appears in each episode
    \item Every-Visit MC: count every time $s$ appears
\end{itemize}

\subsubsection{Why Action Values Matter}
Without a model, state values alone aren't enough. So we need to estimate $q_\pi(s,a)$ to choose actions. But if policy is predictable (deterministic), only sees one action per state, and never learn others. Solution: exploring starts (force random starting state-action pairs) or soft policies (always some probability for all actions).

\subsubsection{MC Control Methods}

Three approaches introduced:

\begin{enumerate}
    \item \textbf{Monte Carlo ES (Exploring Starts)}
    todo
    \item \textbf{On-Policy MC (without exploring starts)}
    todo
    \item \textbf{Off-Policy MC (two policies)}
    todo
    \item 
\end{enumerate}
\subsubsection{Summary}
MC methods learn from experience (no model needed), don't bootstrap (don't use other estimates), simple (just average returns), but must wait until episode ends. Three types: prediction (evaluate policy), on-policy control (learn while exploring), off-policy control (learn optimal while behaving exploratory).

\subsection{Chapter 6}





\section{Terminology}
\begin{enumerate}
    \item Machine Learning (ML)
    \item Reinforcement Learning (RL)
    \item Policy
    \item Agent: The decision-maker who performs actions.
    \item Environment: The world or system in which the agent operates.
    \item State: The situation or condition in which the agent is currently situated.
    \item Action: The possible moves or decisions that the agent can make.
    \item Reward: The feedback or result from the environment based on the agent's actions.
\end{enumerate}


\bibliographystyle{plain}
\bibliography{ref}
\end{document}