



\chapter{Metrics}




\section{seeds}
\paragraph{Seeding Strategy.}
To ensure reproducibility while avoiding correlations induced by sequential seeding,
all experimental runs were initialized from a single \emph{master seed}.
Individual run seeds were deterministically derived using NumPy's
\texttt{SeedSequence} spawning mechanism.
This approach avoids linear relationships between seeds (e.g., $0,1,\dots,14$),
which may introduce inter-stream correlations in certain pseudorandom number
generators
\cite{matsumotoCommonDefectsInitialization2007}.
All randomness sources (environment, NumPy, and PyTorch) were seeded
using the derived run-specific seeds.



\section{intro}
This section defines what data are produced by reinforcement learning training, how we extract
meaningful evaluation scores from that data, and what metrics we compute to compare algorithms.
Because this study uses a small set of $M=3$ environments, \emph{per-environment} results are
the primary evidence, and cross-environment aggregates are treated as descriptive summaries
of the three chosen tasks rather than universal generalizations. \cite{agarwal2021precipice,patterson2024edrl}



\section{Data produced by RL interaction}

During training, at each time step $k$ the agent–environment interaction yields the transition tuple:
\[
(s_k,\,a_k,\,r_k,\,s_{k+1},\,d_k),
\]
where:
\begin{itemize}
  \item $s_k$ is the state (or observation) at time step $k$;
  \item $a_k$ is the action taken by the agent in state $s_k$;
  \item $r_k$ is the scalar reward received after taking action $a_k$;
  \item $s_{k+1}$ is the next state resulting from that action;
  \item $d_k\!\in\!\{0,1\}$ is a terminal indicator: $d_k=1$ if the episode ends after this transition
        (by environment termination or truncation).
\end{itemize}

From this trajectory data, we do \emph{not} make algorithm comparisons directly on training rewards
because training behavior (e.g., exploration noise, replay buffer dynamics, on- vs off-policy sampling)
is algorithm-dependent and not a stable measure of generalization. Instead, we define a fixed
\textbf{evaluation protocol} under which every algorithm is assessed.

\section{Evaluation scores: definitions and notation}

An evaluation episode has an (environment-defined) length $H$, and the \textbf{episodic return}
is defined as:
\[
G \;=\; \sum_{k=0}^{H-1} r_k,
\]
i.e., the accumulated reward over the episode. Although many algorithms optimize discounted returns
internally, we report undiscounted $G$ unless otherwise specified by the benchmark.

We use the following index conventions throughout:

\begin{itemize}
  \item $m\in\{1,2,3\}$: index over evaluated environments (tasks),
  \item $n\in\{1,\dots,N\}$: index over independent training runs (random seeds),
  \item $t$: evaluation checkpoint index associated with training budget (e.g., environment steps),
  \item $T$: the final evaluation checkpoint (end of training budget),
  \item $E$: number of evaluation episodes at each checkpoint.
\end{itemize}

At each checkpoint $t$, we evaluate a frozen policy for $E$ episodes and store the mean
episodic return:
\[
x_{m,n}(t)\;=\;\frac{1}{E}\sum_{e=1}^{E}G^{(e)}_{m,n}(t),
\]
where $G^{(e)}_{m,n}(t)$ is the return of the $e$-th evaluation episode on environment $m$
for seed $n$ at checkpoint $t$.

This $x_{m,n}(t)$ is the \textbf{atomic evaluation score}: all metrics below are computed from
these logged values. At a fixed checkpoint $t$, we can organize these values into a score matrix
\[
X(t)\in\mathbb{R}^{N\times M},\qquad X_{n,m}(t)=x_{m,n}(t),
\]
with rows indexing seeds and columns indexing environments.

\paragraph{Evaluation action rule.}
We fix a single evaluation policy selection rule for all algorithms
(e.g., greedy action for value-based methods; mean/mode action or stochastic sampling for
policy-based methods). This rule must be explicitly documented because it affects reported
performance. \cite{patterson2024edrl}

\section{Uncertainty quantification: 95\% stratified bootstrap}

Reinforcement learning exhibits high variance across independent training runs. To quantify
statistical uncertainty in aggregate summaries, we report \textbf{95\% stratified bootstrap
confidence intervals} for all non-point-estimate metrics. In stratified bootstrap, for each
environment $m$ we resample the set of seeds with replacement to form a bootstrap replicate
of the score matrix; we recompute the statistic on each replicate and take the 2.5\% and
97.5\% percentile bounds. This procedure is recommended in the few-run regime. \cite{agarwal2021precipice}

\section{Per-environment metrics (primary)}

Per-environment metrics operate on the set $\{x_{m,n}(t)\}_{n=1}^N$ for a fixed environment $m$.

\begin{itemize}

\item \textbf{Learning curve (with uncertainty).}  
  \textbf{Definition:} plot the central tendency (mean or median over seeds) of $x_{m,n}(t)$
  against training budget $t$, with pointwise uncertainty bands across seeds.  
  \textbf{Purpose:} reveals the dynamics of learning (speed, plateaus, instability) that a
  single final number cannot capture.

\item \textbf{Final performance at fixed budget $T$ (with CI).}  
  \textbf{Definition:} summarize the set $\{x_{m,n}(T)\}_{n=1}^N$ with a central tendency
  (mean or median) and attach a 95\% stratified bootstrap confidence interval.  
  \textbf{Purpose:} answers ``how good is the agent after the same amount of experience?''
  and avoids incomparable reporting based on best-ever checkpoints. \cite{agarwal2021precipice}

\item \textbf{Sample efficiency.}  
  \textbf{Definition:} one common choice is the discrete area under the curve:
  \[
    \mathrm{AUC}_{m,n}=\sum_{j=1}^{J}x_{m,n}(t_j)\,\bigl(t_j-t_{j-1}\bigr),
  \]
  optionally normalized by the total budget $t_J=T$.  
  \textbf{Purpose:} captures how quickly the agent acquires reward; higher AUC indicates
  faster learning under the same interaction budget.

\item \textbf{Reliability across seeds (variability + tail).}  
  \textbf{Definition:} compute the interquartile range (IQR) of $\{x_{m,n}(T)\}_n$, and, if
  supported by enough seeds, the conditional value at risk (CVaR$_\alpha$) of the worst
  $\alpha$ fraction of seeds.  
  \textbf{Purpose:} captures robustness and worst-case behavior across independent runs.
  \cite{chan2020reliability}

\item \textbf{Success rate (optional).}  
  \textbf{Definition:} if the environment defines a binary success condition, define
  $\mathrm{SR}_{m,n}(t)=\frac{1}{E}\sum_{e=1}^E\mathbb{I}[\text{episode $e$ is success}]$.  
  \textbf{Purpose:} success rate can be more interpretable than raw returns on sparse-goal tasks.

\end{itemize}

\section{Cross-environment metrics (secondary summaries)}

Cross-environment metrics summarize the score matrix $X(T)$ at the final budget. With
$M=3$ diagnostic environments, these are descriptive summaries rather than claims of
general performance across all tasks. \cite{agarwal2021precipice,patterson2024edrl}

\begin{itemize}

\item \textbf{Interquartile mean (IQM) + CI.}  
  \textbf{Definition:} pool the $MN$ values in $X(T)$, trim the lowest 25\% and highest 25\%,
  and average the remaining middle 50\%.  
  \textbf{Purpose:} a robust overall score that discounts extremes and often yields tighter
  uncertainty in the few-run regime. \cite{agarwal2021precipice}

\item \textbf{Aggregate mean and median + CI.}  
  \textbf{Definition:} per-env mean $\bar{x}_m=\frac{1}{N}\sum_{n}x_{m,n}(T)$,
  Mean$=\frac{1}{M}\sum_m\bar{x}_m$, Median$=\mathrm{median}(\bar{x}_1,\dots,\bar{x}_M)$.  
  \textbf{Purpose:} traditional summaries included for comparability; interpret with caution
  when $M$ is small or score scales differ.

\item \textbf{Probability of improvement (pairwise A vs B) + CI.}  
  \textbf{Definition:} for each environment $m$, estimate
  $P_m=\Pr(x^A_{m,\cdot}>x^B_{m,\cdot})$ across seeds; report $\frac{1}{M}\sum_mP_m$ with bootstrap
  uncertainty.  
  \textbf{Purpose:} interpretable probability that A outperforms B on a randomly chosen seed
  and environment. \cite{agarwal2021precipice}

\item \textbf{Performance profile (tail distribution).}  
  \textbf{Definition:} for threshold $\tau$, define
  $\hat{F}(\tau)=\frac{1}{MN}\sum_{m,n}\mathbb{I}[x_{m,n}(T)\ge\tau]$, with pointwise CI bands.  
  \textbf{Purpose:} visualizes the spectrum of outcomes (including failure probabilities).
  \cite{agarwal2021precipice}

\item \textbf{Optimality gap (if a meaningful target).}  
  \textbf{Definition:} for target $\tau$, compute
  $\mathrm{Gap}(\tau)=\tau-\frac{1}{MN}\sum_{m,n}\min(x_{m,n}(T),\tau)$.  
  \textbf{Purpose:} measures shortfall from a predefined target while clipping large outliers.

\end{itemize}

\section{Reporting and reproducibility}

For each experiment we report: number of seeds $N$, evaluation episodes $E$,
total training budget and evaluation frequency, evaluation action rule, and hyperparameter
configuration. To support reproducibility and meaningful comparison, we will provide
code, fixed random seeds, configuration files, and software/hardware details. \cite{henderson2018matters,colas2018seeds}


